---
title: "Investigating NYC Health Inspections and Income"
author: "Alex Pieroni, Ali Gaviser, Marina Chen, Nick Reddy"
date: "20 April 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
subtitle: 'due April 20, 2022 at 11:59 PM'
---

```{r load-packagess, message = FALSE, warning = FALSE, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(fig.width=12, fig.height=4.75)
library(tidyverse) 
library(tidytuesdayR)
library(infer)
library(tidymodels)
```

```{r load-data, warning=FALSE, message = F}
df <- readr::read_csv("https://data.cityofnewyork.us/api/views/43nn-pn8j/rows.csv")
```

# Introduction and Data

## Topic and Motivation

We are interested in this data because food safety is important for a dense
area such as NYC. The data set is huge, allowing freedom of exploration. We
chose health scores and grades because they are widely applicable to each
restaurant and borough (since every restaurant received a grade/score). Although
New York City is considered a homogeneous metropolitan area, there is depth and
variety among, and even within, boroughs. We wanted to highlight these nuances.
Furthermore, it's also important to reveal discrepancies caused by income to
expose inequalities; we wanted to show that socioeconomics can have a
correlation with many facets of life, especially food and health.
  
The study cited below, by Meltzer, et al. (2015), supports the importance
of exploring NYC restaurant data stated above. They found that while health
inspections in restaurants do increase overall grades for all restaurants over
time, benefiting the health of consumers, the result of a health grade can
negatively impact the establishment's immediate sales. For less affluent
restaurant owners or cultural cuisines that may require more preparation, the
restaurant market may suffer in terms of diversity. Thus, the correlation
between income, health grade/score, and borough is important to further
investigate how wealth is intertwined with the location of restaurants and
their scores.
  
## Dataset: NYC Restauraunt Inspections

The data is from the New York Open Data Portal, sourced from TidyTuesday. Only
restaurants in an active role are included. The dataset contains every sustained
violation citation from every full or special program inspection as collected by
the City of NYC Department of Health (www1.nyc.gov/). There are 345,036 obs. of
26 variables; the vast majority were collected from 2015-2022, though two
observations are from the year 1900. The data encompasses a wide variety of
boroughs, scores, cuisine types, and zipcodes. There are both a letter and
number score of the restaurant's hygiene. We will add an income variable from
the 2019 US Census Bureau.
  
Restaurants are rated on a scale starting at 0. A score between and including
0-23 receives an A health grade, 24-27 scores correlate to a B health grade,
and scores of 28 and above indicate a C health grade. A lower score indicates a
better health grade. Therefore, A is best, then B, then lastly, C.

  Sources:

https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-12-11

https://www1.nyc.gov/site/doh/business/food-operators/letter-grading-for-restaurants.page#:~:text=Since%202010%2C%20New%20York%20City,seen%20by%20people%20passing%20by

https://www.census.gov/programs-surveys/acs/news/updates/2019.html

http://www.appam.org/assets/1/7/Impact_paper_9-30-15.pdf

## Research Question:
  What boroughs have the highest and lowest health ratings, and how does that
  correlate to median income?
  
## Hypotheses:

We hypothesize that NYC restaurants in Manhattan, a richer borough, will have 
better health grades. We believe that boroughs with higher median income will 
have higher health grades. Conversely, we expect the Bronx, a borough with a 
lower median income, will have the worst health grades. Overall we expect 
better health grades to be correlated with higher median income and vice versa.

```{r nyc glimpse, message = FALSE, warning = FALSE, include = FALSE}
glimpse(df)
```

# Methodology

```{r new data, message = FALSE, warning = FALSE}
boro_grade_score <- df %>%
  select(BORO, GRADE, SCORE) %>%
  drop_na(BORO, GRADE, SCORE) %>%
  filter(BORO != 0,
         GRADE != 0) %>%
  drop_na(SCORE)
```

```{r boro_grade_score_summarize}
boro_meangrade_income <- boro_grade_score %>%
  select(BORO, GRADE) %>%
  mutate(grade_number = case_when(GRADE == 'A'~ 1, GRADE == 'B'~ 2, GRADE == 'C'~ 3)) %>%
  group_by(BORO) %>%
  drop_na(grade_number) %>%
  summarise(mean_grade = mean(grade_number)) %>%
  arrange(BORO)
boro_meangrade_income["Income"] = c(22232, 31406, 50985, 31992, 39777)
```

```{r boro_meanscore_income}
boro_meanscore_income <- boro_grade_score %>%
  select(BORO, SCORE) %>%
  group_by(BORO) %>%
  drop_na(SCORE) %>%
  summarise(mean_score = mean(SCORE)) %>%
  arrange(BORO)
boro_meanscore_income["Income"] = c(22232, 31406, 50985, 31992, 39777)
```

We made three data sets from our main data. The first one drops the 
non-applicable values in the dataset, and selected the variables of borough, 
grade, and score.

The second data set selects the borough and grade, and organizes each grade with
A = 1, B = 2, and C = 3. From there, we found the 
mean grade number on the same 1-2-3 scale by each borough and added the median 
income from the 2019 US Census Bureau. We dropped non-applicable values.

We only looked at grades A, B, and C, because we found that other investigations
only used these grades, and there are far more data points correlated with these
grades. These are also the grades that the NYC government uses on their website.

The third data set selects the borough and score, dropping non-applicable 
values. The scores are given based on sanitary inspection. A score from 0-13 is 
an A, from 14-27 is a B, and 28 or more points is a C. From there, we found the 
mean score by each borough and added the median income from the 2019 US Census 
Bureau.

```{r figure-1, message = FALSE, warning = FALSE}
boro_grade_score %>%
  arrange(BORO) %>%
  ggplot(mapping = aes(x = BORO, fill = GRADE)) + 
  geom_bar() + 
  labs(title = "FIGURE 1:
Health Grade Letter Ratings in Each NYC Borough",
     x = "Borough",
     y = "Count of Restauraunts")
```

Figure 1 is a histogram of the letter Health Grades for each borough. We used a 
histogram to show the count of each grade while using different colors. This 
allows us to show the number of restaurants in each grade category in each 
borough. This visualization shows that Manhattan has the most amount grade A
restaurants and also has the most amount of total restaurants in the data set 
with a little more than 60,000 observations. On the other hand, Staten Island
has around 5,000 restaurants in this data set. The difference in the amount of
observations for each borough will be a factor worth noting when analyzing our
results.

```{r figure-2, message = FALSE, warning = FALSE}
boro_grade_score %>%
  arrange(BORO) %>%
  ggplot(mapping = aes(x = BORO, fill = GRADE)) + 
  geom_bar(position = "fill") +
  labs(title = "FIGURE 2:
Percent Health Grade Letter Ratings in Each NYC Borough",
     x = "Borough",
     y = "Percent of Restauraunts")
```

Figure 2 is similar to Figure 1, but now shows the frequency of letter 
grades by borough instead by percentages. This figure shows that Staten Island
has the highest percentage of grade A restaurants, which was not seen in Figure
1. This allows us to compare each borough's health ratings relatively, which is 
also important in our an analysis because of the differences in number of 
observations for each borough.

```{r load-packages2, message = FALSE, warning = FALSE, include = FALSE}
knitr::opts_chunk$set(fig.width=12.5, fig.height=9.75) 
```

```{r figure-3, message = FALSE, warning = FALSE}
boro_grade_score %>%
  ggplot(mapping = aes(x = GRADE)) +
  geom_histogram(stat = "count") +
  facet_wrap(~ BORO)+  
  labs(title = "FIGURE 3:
Health Grade faceted by Borough",
       x = "Grade",
       y = "Count")
```

Figure 3 illustrates the frequency of each health letter grade, faceted by 
borough. This would allow us to see the spread of health ratings across each 
borough. It is one method to determine a correlation between restaurant 
health ratings and the restaurant's borough.

```{r figure-4}
boro_grade_score %>%
  filter(SCORE < 40) %>%
ggplot(mapping = aes(x = SCORE))+
  geom_histogram(bins = 20) +
  facet_wrap(~ BORO) +
  geom_vline(aes(color = "A Grade Cutoff",xintercept = 13))+
  geom_vline(aes(color = "B Grade Cutoff",xintercept = 27))+
  labs(title = "FIGURE 4:
Frequency of Scores in All Boroughs",
       y = "Number of Restauraunts",
       x = "Score")+
  scale_color_manual(name = "Cutoff Key", values = c("A Grade Cutoff" = "green", 
                                                  "B Grade Cutoff" = "orange"))
```

Like Figure 3, Figure 4 is a histogram of health ratings faceted by borough.
However, instead of using letter grades, we graphed scores. The faceting allows 
the easy comparison of each borough. We can see that although each borough has a
different frequency of each score, the distribution is quite similar. We chose 
to do scores because it adds more information about restaurant health ratings as
compared to grades. This is because each grade is determined within a range of 
scores. The green vertical lines are at a score of 13, the cutoff for an A 
letter grade. The orange lines indicate the cutoff for a B rating at 27. A 
restaurant with a score of 28 or above is given a C grade.Each borough has a 
peak (of varying size) extremely close to the A/B cutoff point. There are very 
few high quality A ratings. The graph shows that Manhattan has the most A 
ratings, however, it also shows that there are more scores recorded in 
Manhattan in general. However, it's worth noting how much higher the A scores 
are (particularly by the A/B barrier) considering the B scores are not present 
at that much higher of a frequency in comparison to the other boroughs. We 
decided to cutoff the range of scores at 40. We did this because although there 
were scores above 40, proportionally there were not enough scores to actually 
see them in the histogram. Cutting off the C scores at 40 gives each letter 
grade a range of 13, making the graph proportional. 


```{r load-packages3, message = FALSE, warning = FALSE, include = FALSE}
knitr::opts_chunk$set(fig.width=6.25, fig.height=5.5) 
```

```{r figure-5, message = FALSE, warning = FALSE}
ggplot(boro_meangrade_income, mapping = aes(x = Income, y = mean_grade, 
                                            color = BORO)) + 
  geom_point(size = 7) + 
  geom_smooth(method = "lm") + 
  labs(title = "FIGURE 5:
Correlation Between Income and Mean Health Grade",
       subtitle = "Colored by Each Borough",
       y = "Mean Grade (A=1, B=2, C=3)",
       x = "Average Median Income (USD)")
```

Figure 5 above was chosen to clearly show the relationship between mean 
health grades and median income for each borough. A scatter plot is the best 
choice because it clearly shows the relationship between two numerical 
variables, and we can use color to indicate the borough that each point 
represents. The Bronx has the highest mean health grade (closest to a B) and the
lowest median income, while the Manhattan borough has the highest median income 
and the second lowest mean grade (second closest to an A). These correlations
are what we expected, as indicated in our hypothesis. It's worth noting that 
Staten Island has the lowest mean grade by a notable difference, and the second 
highest median income. This could also correlate to what we expected in our 
hypothesis, but it may also be a product of the fact that Staten Island had the 
fewest recorded data/number of restaurants in the data set. 

```{r figure-6, message = FALSE, warning = FALSE}
ggplot(boro_meanscore_income, mapping = aes(x = Income, y = mean_score, 
                                            color = BORO)) + 
  geom_point(size = 7) + 
  geom_smooth(method = "lm") +
  labs(title = "FIGURE 6:
Correlation Between Income and Mean Score",
       subtitle = "Colored by Each Borough",
       y = "Mean Score",
       x = "Average Median Income (USD)")
```
Similarly to Figure 5, Figure 6 correlates mean score with the average median 
income (USD) colored by the borough. This allows us to investigate the question 
using the mean score. The cutoff for an A score is 13, and the range of the mean
scores is from 12.69 to 13.83.

```{r mean grade, message = FALSE, warning = FALSE}
df %>%
  mutate(grade_number = case_when(GRADE == 'A'~ 1, GRADE == 'B'~ 2, GRADE == 'C'~ 3)) %>%
  drop_na(grade_number) %>%
  summarise(mean_grade = mean(grade_number))
```
The mean grade of restaurants across the five boroughs is between an A and a B. 
The mean grade is 1.22, with A being 1 and B being 2; therefore, the mean grade 
is actually closer to an A than a B.

```{r mean grade borough, message = FALSE, warning = FALSE}
boro_meangrade_income %>%
  arrange(desc(mean_grade)) 
```

While the mean for all restaurants across all boroughs is 1.22, the mean grade 
in Manhattan is 1.21. This is slightly lower than the total mean, leading us to 
believe that restaurants in Manhattan have a higher grade on average. In other 
words, in comparison to all of New York City, the Manhattan restaurants have a 
mean grade (in this sample) that is slightly closer to an A grade.

As mentioned above, the mean for all restaurants across all boroughs is 1.22. 
The mean grade in the Bronx, however, is 1.27. This is higher than both the mean
for New York City restaurants, and it is actually the highest mean grade out of 
each individual borough. Therefore, the Bronx has a mean grade that is closer to
B than any other borough. The health ratings in this dataset are lowest for the 
Bronx out of all five boroughs. 

```{r mean score borough, message = FALSE, warning = FALSE}
boro_meanscore_income %>%
  arrange(desc(mean_score)) 
```

Similarly to the results from our mean score, our summary results tell us that 
the Bronx has the highest (and therefore least healthy) mean score, while 
Staten Island has the lowest mean score. This borough also has the fewer data
points when compared to Manhattan.

# Results

To answer what boroughs have the highest and lowest health ratings, we will find
summary statistics for each boro.

## Null Hypothesis 1

Null Hypothesis: The mean health score of Manhattan restaurants is equal to the 
mean of all other boroughs' restaurant scores.

Alternative Hypothesis: The mean health score of Manhattan restaurants is less
than the mean of all other boroughs' restaurant scores.

*Note: the closer the score is to 0, the less violations and more up to code a
restaurant is*

M: Manhattan

O: Other Boroughs

 $\\H_0: \mu_M = \mu_O$
 
 $\\H_1: \mu_M < \mu_O$

```{r load-packages4, message = FALSE, warning = FALSE, include = FALSE}
knitr::opts_chunk$set(fig.width=12.5, fig.height=7.5) 
```


```{r p test Manhattan, message = FALSE, warning = FALSE}
set.seed(212)
null_dist_data <- boro_grade_score %>%
  mutate(Manhattan = ifelse(BORO == "Manhattan", "Manhattan", "Not Manhattan"))
null_dist <- null_dist_data %>% 
  specify(response = SCORE, explanatory = Manhattan) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 100, type = "permute") %>%
  calculate(stat = "diff in means", order = c("Manhattan", "Not Manhattan"))
obs_sta_Manhattan <- boro_grade_score  %>%
  filter(BORO == "Manhattan") %>%
  summarise(mean_score_M = mean(SCORE)) %>%
  pull()
obs_sta_NotManhattan <- boro_grade_score %>%
  filter(BORO != "Manhattan") %>%
  summarise(mean_score_NM = mean(SCORE)) %>%
  pull()
obs_sta_M_NM = obs_sta_Manhattan - obs_sta_NotManhattan
visualize(null_dist) +
  shade_p_value(obs_stat = obs_sta_M_NM, direction = "less") +
  labs(x = "Sample Difference", y = "Count")
null_dist %>%
  get_p_value(obs_stat = obs_sta_M_NM, direction = "less")
```

The p-value for this simulation is 0, so $p < \alpha$. Since the p-value is less
  than the significance level, we reject the null hypothesis that the mean 
  health score of Manhattan restaurants is equal to the mean of all other 
  boroughs' restaurant score. Instead, we favor of the alternative hypothesis 
  that the mean health score of Manhattan restaurants is less than the mean of 
  all other boroughs' restaurant scores. We used 100 reps because it was
  sufficient for our data set and our computers ran out of memory with anything
  over 100.
  
Now we will perform hypothesis testing for the Bronx.

## Null Hypothesis 2

Null Hypothesis: The mean health score of Bronx restaurants is equal to the 
mean of all other boroughs' restaurant scores.

Alternative Hypothesis: The mean health score of Bronx restaurants is greater
than the mean of all other boroughs' restaurant scores.

*Note: the closer the score is to 0, the less violations and more up to code a
restaurant is*

B: Bronx

O: Other Boroughs

 $\\H_0: \mu_B = \mu_O$
 
 $\\H_1: \mu_B > \mu_O$

```{r p test Bronx, message = FALSE, warning = FALSE}
set.seed(212)
null_dist_data_2 <- boro_grade_score %>%
  mutate(Bronx = ifelse(BORO == "Bronx", "Bronx", "Not Bronx"))
null_dist_2 <- null_dist_data_2 %>% 
  specify(response = SCORE, explanatory = Bronx) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 100, type = "permute") %>%
  calculate(stat = "diff in means", order = c("Bronx", "Not Bronx"))
obs_sta_Bronx <- boro_grade_score  %>%
  filter(BORO == "Bronx") %>%
  summarise(mean_score_B = mean(SCORE)) %>%
  pull()
obs_sta_NotBronx <- boro_grade_score %>%
  filter(BORO != "Bronx") %>%
  summarise(mean_score_NB = mean(SCORE)) %>%
  pull()
obs_sta_B_NB = obs_sta_Bronx - obs_sta_NotBronx
visualize(null_dist) +
  shade_p_value(obs_stat = obs_sta_B_NB, direction = "greater") +
  labs(x = "Sample Difference", y = "Count")
null_dist %>%
  get_p_value(obs_stat = obs_sta_B_NB, direction = "greater")
```

The p-value for this simulation is also 0, so $p < \alpha$. Since the p-value
  is less than the significance level, we reject the null hypothesis that the
  mean health score of Bronx restaurants is equal to the mean of all other 
  boroughs' restaurant score. Instead, we favor of the alternative hypothesis 
  that the mean health score of Bronx restaurants is greater than the mean of 
  all other boroughs' restaurant scores. 

## Linear Regression

```{r regression, message = FALSE, warning = FALSE}
regression <- linear_reg() %>%
  set_engine("lm") %>%
  fit(mean_grade ~ Income, data = boro_meangrade_income)
tidy(regression)
```
For our linear regression model, we are trying to predict the mean grade of a
borough from the median income of the borough. Interpreting the intercept, for a 
borough with a median income of 0 dollars, on average the mean grade is 1.3, 
which in this context does not make sense because a borough will never have a 
median income of 0 dollars. Interpreting the slope, for an increase in median
income by 1 dollar, on average the mean grade is expected to decrease by 
.00000212 dollars. However, for an alpha of .05, the p-value for this predictor, 
.183, is not significant, therefore there is not enough evidence to conclude
a relationship exists between median income and mean grade and are not 
necessarily correlated. 

We compared the mean health score for restaurants in Manhattan to the mean 
health score for restaurants in the other four boroughs. Our data in the 
methodology section shows that Manhattan has the highest income of the five 
boroughs. 
 

# Discussion

## Conclusions:

The boroughs that have the highest and lowest health ratings were Manhattan and
the Bronx, respectively. The mean grade for all of New York City's boroughs
collectively was 1.22. Manhattan's was 1.21, which is the second closest to an 
A in comparison to all of New York City. Staten Island had a mean grade of 1.18, 
the lowest (and best) health grade of the five boroughs. Staten Island also has 
the second highest median income. On the other hand, the Bronx had a mean
grade of 1.27, which is closer to a B than any other borough. The same trend 
holds for mean health scores. Our overall hypothesis was also supported: income 
was indeed correlated with which borough had the highest mean health grade, and 
vice versa. However, these correlations are just summary statistics of the data 
set. Our hypothesis test is more informative. The hypothesis test for Manhattan 
having a lower mean score than the other boroughs yielded a significant p-value 
of 0 at the $\alpha = 0.05$ significance level. The hypothesis test for the 
Bronx having a higher mean score than the other boroughs yielded a significant 
p-value of 0 with $\alpha = 0.05$. Therefore we were able to reject the null 
hypothesis in both cases.

## Limitations:

We rejected the null hypothesis both of our hypothesis tests because the 
p-values were significant, but when we ran our linear regression, the p-value
was not significant. This means we should be hesitant about the validity of our 
null hypothesis rejection, warranting further investigation into our research 
topic. Our hypothesis test also used 1,000 repetitions; 10,000 reps were too 
high for a data set of this size, the test wouldn't run. This may have impacted 
our results as well, but it was an unavoidable choice. 

In terms of the data, it was collected between 2015 and 2022, with a couple 
entries actually collected in 1900. The health ratings could have drastically 
changed over this time period. This could have potentially increased or 
decreased the correlations we observed. Moreover, while data sets with a lot of 
information can be helpful, however our data set had so much information that it 
was almost too much to work with. A lot of the restaurants had NA values, some 
of the scores didn't make sense, etc. This forced us to filter out some data 
points in order to produce figures and data that we could actually analyze. 
Cutting the data made the most sense, however, it could have impacted our 
results. Lastly, although out of our control, we don't know how health 
ratings are given or biases that might go into it. Some health grades may not be 
deserved for one reason or another. 

## Further Research: 
In the future we would like to compare health grades by smaller divisions than 
borough, perhaps going into New York City neighborhoods as well as comparing 
scores to the neighborhood's mean incomes. Additionally, we would like to add in 
another variable: cuisine. It would be interesting to see how health ratings 
change by cuisine in general, as well as across boroughs. Perhaps the racial and 
ethnic makeup of a borough may impact the ratings of a particular group's 
cuisine. 

In terms of further developing our findings, in the future we would like to try 
to assess the true mean health scores with confidence interval within each 
borough and New York as a whole, and then compare that to the median income 
within each borough. This would be a further development since we just assessed 
the health grades of Manhattan and the Bronx relative to the other boroughs. 